No need to know: astrophysics-agnostic gravitational wave cosmology

### 2024-03-10 src/tex/ms.tex
[showyourwork] overleaf sync

\subsection{Projections for Future Measurements}
This result implies that we can expect to measure \Ho{} to \variable{output/nonparh0percent.txt}\% by the end of \ac{O5} using the method presented in this work.
However, by the time of \ac{O5}, the \ac{GW} detector network is projected to detect \acp{BBH} up to redshift $\sim 3$, with most sources lying near redshift $\sim 1.2$ \citep{chen_distance_2021}.
Additionally, next-generation detectors will be sensitive to sources up to redshift $\sim 40$.
This means that future observations of \acp{GW} will be more sensitive $H(z>1)$ than to \Ho, and can therefore constrain several additional cosmological parameters.
We demonstrate this by repeating the same analysis while also fitting for the local matter density, \Omm.
The result is shown in Figure~\ref{fig:O5_corner}.
We find \ac{O5} observations to be most sensitive to \result{$H(z=X)$}, which results in similarly informative constraints on \Omm{} as on \Ho.
\begin{itemize}
    \item There is a strong degeneracy in these two cosmological parameters, caused by the sensitivity at z=1
    \item explain why - balance between slope and intercept.
\end{itemize}

@@

\begin{itemize}
    \item We are able to get constraints on cosmological parameters with a very flexible model for the mass distribution. 
    \item The success of the flexible model in recovering an informative and accurate posterior on \Ho{} demonstrates that there is no need to know the \ac{CBC} mass distribution's morphology in order to constrain cosmological parameters with it.
    \item This is because the information in spectral sirens comes from assuming events come from the same overall mass distribution across redshift, and that mass distribution doesn't evolve with redshift in the same exact way as would be mimicked by cosmological redshifting. This assumption is very basic, not an assumption that we know the exact astrophyisics of \ac{BBH} formation (i.e. PPISN feature location).
    \item We have also predicted a \variable{output/nonparh0percent.txt}\% measurement of \Ho{} by the end of \ac{O5} using non-parametric spectral sirens.
A similar study with parametric methods \citep{Chen:2024gdn} predicts an \result{X\%} measurement, demonstrating a \result{negligible/substantial/marginal} increase in precision.
However, the true population of compact objects in the Universe is unknown, so it is not possible to know the true functional form of the mass distribution, so systematic uncertainties arising from an incorrect choice for the functional form of the mass distribution are therefore inevitable.
With current detectors, these systematic effects are dwarfed by statistical uncertainties.
However, next-generation detectors will herald in enough \ac{GW} observations to substantially decrease statistical uncertainty in these measurements: \citet{pierra_study_2023} show that incorrect assumptions about the shape of the mass distribution can lead to \result{$\sim3\sigma$} systematic biases in \Ho when catalogs contain 2,000 events.
Thus, the additional statistical uncertainty introduced by non-parametric approaches may be preferable to the systematic errors associated with choosing a parametric model.
    \item additionally, in \ac{O5} we will be most sensitive to \result{$z=X$}. This is larger than we expect to see EM counterparts to, and larger than we expect galaxy surveys to be complete to (is this true?) so upgrades to current facilities will allow spectral sirens to probe H(z) to relatively unexplored redshifts. This is not unique to our nonparametric method, but to spectral sirens in general.
    \item future/in prep. work will do the same thing but for a mass distribution that is allowed to evolve with redshift, similarly to the parametric analysis done in~\cite{ezquiaga_spectral_2022}, but with the evolution with redshift allowed to be arbitrary so long as it does not mimic cosmology. Because of ~\citet{ezquiaga_spectral_2022}, we think we could get a useful measurement if we include evolving peak shapes/positions in the mass distribution, provided there are multiple features present (which we know there are) and that they do not evolve in the same exact way as cosmology would - i.e. all features do not monotonically move to higher source frame mass with increasing redshift, all at the same rate. That scenario would be astrophysically unlikely, especially since the processes changing the features are all distinct from one another.
    \item GPs are good bc they can do correlations/generalize to multiple dimensions naturally. Our choice of kernel allows us to do this, but its not possible with the autoregressive kernel.
    \item in XG we will get most bang for our buck with spectral sirens
\end{itemize}
\begin{acknowledgments}
    The authors thank Reed Essick, Utkarsh Mali, Ben Farr, Maya Fishbach,  for helpful conversations.
\end{acknowledgments}

@@

After passing the simulated events through projected detector selection effects, the resulting catalog has \variable{output/num_found_events.txt}events, including \result{X} binary neutron star systems, \result{Y} NSBH systems, and \result{Z} \ac{BBH} systems, consistent with the numbers projected for \ac{O5} by \citet{kiendrebogo_observing_2023}.
We use the software package \texttt{GWMockCat} \citep{farah_things_2023} to simulate posterior samples for these events with measurement uncertainties typical of those expected from \ac{O5} detectors.
\texttt{GWMockCat} also simulates a set of software injections, which we use to estimate selection effects in the inference.
To determine the detectability of both injections and simulated events in O5, we use the projected \ac{O5} LIGO power spectral density \citep{dccpage} to calculate observed signal-to-noise ratios $\rho_{\text{obs}}$, and we consider events and injections with $\rho_{\text{obs}}>8$ to be detectable. 
The full procedure for this mock data generation process is described in \citet{fishbach_where, farah_things_2023, essick_dagnabbit_2023}, and the exact settings used for \texttt{GWMockCat} are made available in the accompanying data release. \comment{TODO:put a showyourwork link here.}

>>>>

\subsection{Projections for Future Measurements}
This result implies that we can expect to measure \Ho{} to \variable{output/nonparh0percent.txt}\% by the end of \ac{O5} using the method presented in this work.
However, by the time of \ac{O5}, the \ac{GW} detector network is projected to detect \acp{BBH} up to redshift $\sim 3$, with most sources lying near redshift $\sim 1.2$ \citep{chen_distance_2021}.
Additionally, next-generation detectors will be sensitive to sources up to redshift $\sim 40$.
This means that future observations of \acp{GW} will be more sensitive $H(z>1)$ than to \Ho, and can therefore constrain several additional cosmological parameters.
We demonstrate this by repeating the same analysis while also fitting for the local matter density, \Omm.
The result is shown in Figure~\ref{fig:O5_corner}.
We find \ac{O5} observations to be most sensitive to \result{$H(z=\variable{output/mostsensitivez.txt})$}, which results in similarly informative constraints on \Omm{} as on \Ho.
\begin{itemize}
    \item There is a strong degeneracy in these two cosmological parameters, caused by the sensitivity at z=1
    \item explain why - balance between slope and intercept.
\end{itemize}

@@

\begin{itemize}
    \item We are able to get constraints on cosmological parameters with a very flexible model for the mass distribution. 
    \item The success of the flexible model in recovering an informative and accurate posterior on \Ho{} demonstrates that there is no need to know the \ac{CBC} mass distribution's morphology in order to constrain cosmological parameters with it.
    \item This is because the information in spectral sirens comes from assuming events come from the same overall mass distribution across redshift, and that mass distribution doesn't evolve with redshift in the same exact way as would be mimicked by cosmological redshifting. This assumption is very basic, not an assumption that we know the exact astrophyisics of \ac{BBH} formation (i.e. PPISN feature location).
    \item We have also predicted a \variable{output/nonparh0percent.txt}\% measurement of \Ho{} by the end of \ac{O5} using non-parametric spectral sirens.
A similar study with parametric methods \citep{Chen:2024gdn} predicts an \result{X\%} measurement, demonstrating a \result{negligible/substantial/marginal} increase in precision.
However, the true population of compact objects in the Universe is unknown, so it is not possible to know the true functional form of the mass distribution, so systematic uncertainties arising from an incorrect choice for the functional form of the mass distribution are therefore inevitable.
With current detectors, these systematic effects are dwarfed by statistical uncertainties.
However, next-generation detectors will herald in enough \ac{GW} observations to substantially decrease statistical uncertainty in these measurements: \citet{pierra_study_2023} show that incorrect assumptions about the shape of the mass distribution can lead to \result{$\sim3\sigma$} systematic biases in \Ho when catalogs contain 2,000 events.
Thus, the additional statistical uncertainty introduced by non-parametric approaches may be preferable to the systematic errors associated with choosing a parametric model.
    \item additionally, in \ac{O5} we will be most sensitive to $z=\variable{output/mostsensitivez.txt}$. This is larger than we expect to see EM counterparts to, and larger than we expect galaxy surveys to be complete to (is this true?) so upgrades to current facilities will allow spectral sirens to probe H(z) to relatively unexplored redshifts. This is not unique to our nonparametric method, but to spectral sirens in general.
    \item future/in prep. work will do the same thing but for a mass distribution that is allowed to evolve with redshift, similarly to the parametric analysis done in~\cite{ezquiaga_spectral_2022}, but with the evolution with redshift allowed to be arbitrary so long as it does not mimic cosmology. Because of ~\citet{ezquiaga_spectral_2022}, we think we could get a useful measurement if we include evolving peak shapes/positions in the mass distribution, provided there are multiple features present (which we know there are) and that they do not evolve in the same exact way as cosmology would - i.e. all features do not monotonically move to higher source frame mass with increasing redshift, all at the same rate. That scenario would be astrophysically unlikely, especially since the processes changing the features are all distinct from one another.
    \item GPs are good bc they can do correlations/generalize to multiple dimensions naturally. Our choice of kernel allows us to do this, but its not possible with the autoregressive kernel.
    \item in XG we will get most bang for our buck with spectral sirens
\end{itemize}
\begin{acknowledgments}
    The authors thank Reed Essick, Utkarsh Mali, Ben Farr, Maya Fishbach,  for helpful conversations.
\end{acknowledgments}

@@

After passing the simulated events through projected detector selection effects, the resulting catalog has \variable{output/num_found_events.txt}events, %including \result{X} binary neutron star systems, \result{Y} NSBH systems, and \result{Z} \ac{BBH} systems, 
consistent with the numbers projected for \ac{O5} by \citet{kiendrebogo_observing_2023}.
We use the software package \texttt{GWMockCat} \citep{farah_things_2023} to simulate posterior samples for these events with measurement uncertainties typical of those expected from \ac{O5} detectors.
\texttt{GWMockCat} also simulates a set of software injections, which we use to estimate selection effects in the inference.
To determine the detectability of both injections and simulated events in O5, we use the projected \ac{O5} LIGO power spectral density \citep{dccpage} to calculate observed signal-to-noise ratios $\rho_{\text{obs}}$, and we consider events and injections with $\rho_{\text{obs}}>8$ to be detectable. 
The full procedure for this mock data generation process is described in \citet{fishbach_where, farah_things_2023, essick_dagnabbit_2023}, and the exact settings used for \texttt{GWMockCat} are made available in the accompanying data release. \comment{TODO:put a showyourwork link here.}
